---
title: "Shifting in Newspaper Presidential Endorsements Over Time"
author: "Gideon Leshner"
description: "An analysis of whether U.S. media (newspapers in this case) have become more politically consistent across presidential elections"
output:
  distill::distill_article:
    self_contained: false
---

# Introduction

I'll look at trends of media political behavior over time using the "endorsements dataset", which includes newspaper presidential endorsements from different election years. My research question is: Do newspapers always endorse the same parties across all elections? My hypothesis is that newspapers have been more consistent in their endorsements because they are motivated to maintain stable political identities rather than changing parties frequently due to growing political polarization and higher audience expectations. In order to determine whether endorsement consistency varies over time, the primary explanatory variable is 'election year', which is defined as the year of each presidential election. In order to quantify the variation in media endorsement consistency across elections, I will also create a variable that indicates whether a newspaper endorsed the same party across all elections in the dataset. The outcome variable is 'endorsement consistency', which is a binary indicator that is equal to 1 if a newspaper always endorsed the same party across the dataset and 0 if it switched parties at least once. A positive correlation between election year and endorsement consistency would indicate that newspapers appearing in later election years are more likely to belong to the always-consistent group. If there is no trend, the hypothesis would be refuted.

# Data

### Load the Data

This project uses a cross-sectional time-series design because it compares newspaper endorsements across different publications and election years.

```{r, echo=FALSE}
library(tidyverse)
endorsements <- read_csv("endorsements.csv")
glimpse(endorsements)
```

```{r, echo=FALSE}
endorsements <- mutate(endorsements, 
               party_of_candidate = 
                 case_when(endorsed == "Obama" | endorsed == "Biden" | 
                    endorsed == "Clinton" | endorsed == "Gore" | 
                    endorsed == "Kerry" ~ "Democrat",
                    endorsed == "Trump" | endorsed == "Romney" | 
                    endorsed == "McCain" | endorsed == "Bush" | 
                    endorsed == "Dole" ~ "Republican"))
endorsements <- select(endorsements, publication, 
                       year, endorsed, party_of_candidate)
endorsements <- na.omit(endorsements)
```

### Simple Data Visualization (Figure 1)

```{r, echo = FALSE}
ggplot(endorsements, aes(x = factor(year), fill = party_of_candidate)) +
  geom_bar(position = "dodge") +
  labs(title = "Endorsements by Year",
       x = "Year",
       y = "Count",
       fill = "Political Party")
```

Figure 1:
This figure shows the number of newspaper endorsements for Democratic and Republican candidates in each election year. Democratic endorsements generally increase over time, while Republican endorsements appear to decline in more recent elections. This suggests a shift in overall endorsement patterns across years.

# Results

### Always-Consistent Table

```{r, echo=FALSE}
endorsements <- endorsements |>
  arrange(publication, year) |>
  group_by(publication) |>
  mutate(prop_dem = mean(party_of_candidate == "Democrat")) |>
  mutate(consistency_status = if_else(prop_dem == 1 | prop_dem == 0, "Consistent", "Has Switched")) |>
  mutate(consistent = if_else(consistency_status == "Consistent", 1, 0)) |>
  ungroup() 

consistency_trends <- endorsements |>
  filter(consistency_status == "Consistent") |>
  distinct(publication, year) |>
  group_by(year) |>
  summarize(consistent_count = n())

library(knitr)
consistency_trends |>
  kable(
    col.names = c("Election Year", "Number of Always-Consistent Publications"),
    digits = 0)
```

### Data Visualization (Figure 2)

```{r, echo=FALSE}
ggplot(consistency_trends, aes(x = year, y = consistent_count)) +
  geom_line() +
  geom_point() +
  labs(title = "Trend of Always-Consistent Newspaper Endorsements",
       x = "Election Year",
       y = "Number of Always-Consistent Publications")
```

Figure 2:
This figure displays the number of newspapers that consistently endorsed the same party across all election years. The number of consistent newspapers peaks around 2000 and then declines over time. This indicates that fewer newspapers maintain the same political alignment in recent elections.

### Analysis

Consistency, in this dataset, is when a newspaper never switches political parties across all of the election cycles recorded in the dataset. Under this logic, a publication is labeled "Consistent" only if it has a 100% track record of endorsing either exclusively Democrats or Republicans. The data visualization indicates that the total number of consistent newspaper endorsements peaked around the year 2000 and has since had a significant decline. My hypothesis was that growing political polarization would lead to higher endorsement consistency, but the results actually show the opposite trend occurring recently. By 2020, the number of publications maintaining a consistent endorsement dropped to its lowest point since 1988, falling below 20 publications. This suggests that newspapers are either increasingly willing to switch their support or, more likely given the current media environment, they are no longer willing to make formal presidential endorsements. Therefore, the evidence does not support the theory that media outlets are maintaining stable political identities. Instead, the normally traditional model of consistent endorsements appears to be weakening in the last couple of years.

### Regression Table

```{r, echo=FALSE}
library(broom)
library(knitr)

model <- lm(consistent ~ year, data = endorsements)

tidy(model) |>
  mutate(term = if_else(term == "year", "Year", "Intercept")) |>
  kable(
    digits = 3,
    col.names = c("Term", "Estimate", "Std. Error", "t value", "p value")
  )
```

### Interpretation

The regression model shows a small positive relationship between election year and endorsement consistency, with a coefficient of about 0.003, meaning that observations in later election years are associated with roughly a 0.3 percentage point increase in the probability that a newspaper is in the always-consistent group. However, the effect is very small substantively. Finally, the p-value for the Year coefficient is 0.104, which is greater than 0.05, indicating that the relationship is not statistically significant. This means there is no strong evidence that endorsement consistency has changed over time, and the observed trend could be due to random chance. This relationship should not be interpreted as causal because the model does not control for other factors that could affect newspaper endorsements and their consistency, such as ownership or audience.

# Conclusion

### Final Conclusion

Overall, the results do not support my hypothesis that newspapers have become more consistent in their endorsements over time. While the estimated relationship is slightly positive, it is not statistically significant and is substantively very small. One limitation of this analysis is that the measure of consistency is simplified and does not capture newspapers that stop making endorsements altogether. Additionally, the model only includes time as an explanatory variable, ignoring other factors such as newspaper ownership, audience, or regional political trends. These limitations can threaten inference and conclusions because confounding or not calculated variables could influence endorsement behavior. The analysis could be improved by collecting more detailed data on newspaper characteristics and modeling changes in endorsements more precisely over time.

# AI Disclosure

I used AI tools to help clarify that my research question was aligned with the tables and visualizations I was making (it wasn't). I was producing results and not clarifying my always-consistent groups category. I was talking about previous elections and their consistency when I wanted to talk about all elections as a whole, if a newspaper was always consistent.

I also used AI tools to help create labels for my regression table and making it a nicely formatted table.

# Code

```{r printo-code, ref.label = knitr::all_labels(), echo=TRUE, eval= FALSE}
```
