---
title: "Shifts in Newspaper Presidential Endorsements Over Time"
author: "Gideon Leshner"
description: "An analysis of whether U.S newspapers have become more politically consistent across presidential elections, 1988-2020"
output:
  distill::distill_article:
    self_contained: false
---

# Introduction

In this project we will look at the trends of newspapers and political behavior over time using the "endorsements dataset". This includes newspaper presidential endorsements from different election years. My research question for this project is: Do newspapers always endorse the same parties across all elections? My hypothesis is that newspapers have stayed consistent in their endorsements because they aim to maintain stable political identities, especially as political polarization increases and audience expectations grow. 

In order to determine whether endorsement consistency varies over time, the primary explanatory variable is 'election year', which is defined as the year of each presidential election. In order to quantify media endorsement consistency across elections, I will also create a variable that indicates whether a newspaper endorsed the same party across all elections in the dataset. The outcome variable is 'endorsement consistency', that is equal to 1 if a newspaper always endorsed the same party across the dataset and 0 if it switched parties at least once. A positive correlation between election year and endorsement consistency would indicate that newspapers appearing in later election years are more likely to belong to the always-consistent group. If there is no trend, the hypothesis would be refuted.

# Data

### Load the Data

This project uses a cross-sectional design. The main outcome variable is endorsement consistency, which equals 1 if a newspaper endorsed only one political party across all of its recorded elections and 0 if it switched parties at least once. My hypothesis would be supported if endorsement consistency increased in later election years.


```{r, echo=FALSE}
library(tidyverse)
endorsements <- read_csv("endorsements.csv")
```
```{r, echo=FALSE}
endorsements <- mutate(endorsements, 
  party_of_candidate = case_when(endorsed == "Trump" | endorsed == "Romney" | 
                                 endorsed == "McCain" | endorsed == "Bush" | 
                                 endorsed == "Dole" ~ "Republican",
                                 endorsed == "Obama" | endorsed == "Biden" | 
                                 endorsed == "Clinton" | endorsed == "Gore" | 
                                 endorsed == "Kerry" ~ "Democrat"))
endorsements <- select(endorsements, publication, 
                       year, endorsed, party_of_candidate)
endorsements <- na.omit(endorsements)
```

### Data Visualization (Figure 1)

```{r, echo = FALSE}
ggplot(endorsements, aes(x = factor(year), fill = party_of_candidate)) +
  geom_bar(position = "dodge") +
  labs(title = "Endorsements by Year",
       x = "Year",
       y = "Count",
       fill = "Political Party")+ 
    scale_fill_manual(values = c("Democrat" = "blue", "Republican" = "red"))
```

Figure 1:
This figure shows the number of newspaper endorsements for Democratic and Republican candidates in each election year. Democratic endorsements generally increase over time, while Republican endorsements appear to decline in more recent elections. This suggests a shift in overall endorsement patterns across years.

### Data Visualization (Figure 2)
```{r, echo=FALSE}
endorsements <- endorsements |>
  arrange(publication, year) |>
  group_by(publication) |>
  mutate(prop_dem = mean(party_of_candidate == "Democrat")) |>
  mutate(consistency_status = if_else(prop_dem == 1 | prop_dem == 0, "Consistent", "Has Switched")) |>
  mutate(consistent = if_else(consistency_status == "Consistent", 1, 0)) |>
  ungroup()

endorsements |>
  count(consistent) |>
  ggplot(aes(x = factor(consistent), y = n)) +
  geom_col() +
  labs(
    x = "Consistency",
    y = "Number of Newspapers",
    title = "Newspaper Endorsement Consistency")
```
Figure 2: 
This figure shows the number of newspapers that consistently endorsed the same party across all elections compared to those that switched at least once. The majority of newspapers are not fully consistent, indicating that party switching is relatively common.

# Results

### Always-Consistent Table

```{r, echo=FALSE}
consistency_trends <- endorsements |>
  filter(consistency_status == "Consistent") |>
  distinct(publication, year) |>
  group_by(year) |>
  summarize(consistent_count = n())

library(knitr)
consistency_trends |>
  kable(
    col.names = c("Election Year", "Number of Always-Consistent Publications"),
    digits = 0)
```

### Data Visualization (Figure 3)

```{r, echo=FALSE}
ggplot(consistency_trends, aes(x = year, y = consistent_count)) +
  geom_line() +
  geom_point() +
  labs(title = "Trend of Always-Consistent Newspaper Endorsements",
       x = "Election Year",
       y = "Number of Always-Consistent Publications")
```

Figure 3:
This figure displays the number of newspapers that consistently endorsed the same party across all election years. The number of consistent newspapers peaks around 2000 and then declines over time. This indicates that fewer newspapers maintain the same political alignment in recent elections. Although Figure 3 shows a decline in always-consistent newspapers after 2000, the regression tests to follow will highlight the overall relationship between election year and consistency across the dataset. 

### Analysis

Consistency, in this dataset, is when a newspaper never switches political parties across all of the election cycles recorded in the dataset. Under this logic, a publication is labeled "Consistent" only if it has a 100% track record of endorsing either exclusively Democrats or Republicans. The data visualization indicates that the total number of consistent newspaper endorsements peaked around the year 2000 and has since had a significant decline. My hypothesis was that growing political polarization would lead to higher endorsement consistency, but the results actually show the opposite trend occurring recently. By 2020, the number of publications maintaining a consistent endorsement dropped to its lowest point since 1988, falling below 20 publications. This suggests that newspapers are either increasingly willing to switch their support or, more likely given the current media environment, they are no longer willing to make formal presidential endorsements. Therefore, the evidence does not support the theory that media outlets are maintaining stable political identities. Instead, the normally traditional model of consistent endorsements appears to be weakening in the last couple of years.

### Regression Table

```{r, echo=FALSE}
library(broom)
library(knitr)

reg_1 <- lm(consistent ~ year, data = endorsements)

tidy(reg_1) |>
  mutate(term = if_else(term == "year", "Election Year", "Intercept")) |>
  kable(
    digits = 3,
    col.names = c("Term", "Coefficient", "Standard Error", "t value", "p value")
  )
```

### Interpretation

The regression model shows a small positive relationship between election year and endorsement consistency, with a coefficient of about 0.003, meaning that observations in later election years are associated with roughly a 0.3 percentage point increase in the probability that a newspaper is in the always-consistent group. However, the effect is very small substantively. Finally, the p-value for the Year coefficient is 0.104, which is greater than 0.05, indicating that the relationship is not statistically significant. This means there is no strong evidence that endorsement consistency has changed over time, and the observed trend could be due to random chance. This relationship should not be interpreted as causal because the model does not control for other factors that could affect newspaper endorsements and their consistency, such as ownership or audience.

# Conclusion

### Final Conclusion

Overall, the results do not support my hypothesis that newspapers have become more consistent in their endorsements over time. While the estimated relationship is slightly positive, it is not statistically significant and is substantively very small. One limitation of this analysis is that the measure of consistency is simplified and does not capture newspapers that stop making endorsements altogether. Another limitation is that newspapers with fewer recorded election years may appear more consistent simply because they had fewer chances to switch parties. Additionally, the model only includes time as an explanatory variable, ignoring other factors such as newspaper ownership, audience, or regional political trends. These limitations can threaten inference and conclusions because confounding or not calculated variables could influence endorsement behavior. The analysis could be improved by collecting more detailed data on newspaper characteristics and modeling changes in endorsements more precisely over time.

# AI Disclosure
I used AI tools to help create labels for my regression table and making it a nicely formatted table.

I also used AI tools to see if my labels were misleading, leading the viewer to perceive the visualizations were something they werent

# Code

```{r printo-code, ref.label = knitr::all_labels(), echo=TRUE, eval= FALSE}
```
